---
title: "Random Forest Challenge"
subtitle: "The Power of Weak Learners"
repo: https://github.com/haileypease/randomForestChallenge
format:
  html:
    theme: default
    toc: true
    code-fold: false
    fig-width: 10
    fig-height: 6
  pdf:
    documentclass: article
    fig-width: 10
    fig-height: 6
    keep-tex: false
execute:
  echo: false
  eval: true
---

# 🌲 Random Forest Challenge - The Power of Weak Learners

```{r}
#| label: load-and-model-r
#| echo: false
#| message: false
#| warning: false

# Load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(randomForest))

# Load data
sales_data <- read.csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Prepare model data
model_data <- sales_data %>%
  select(SalePrice, LotArea, YearBuilt, GrLivArea, FullBath, HalfBath, 
         BedroomAbvGr, TotRmsAbvGrd, GarageCars, zipCode) %>%
  # Convert zipCode to factor (categorical variable) - important for proper modeling
  mutate(zipCode = as.factor(zipCode)) %>%
  na.omit()

cat("Data prepared with zipCode as categorical variable\n")
cat("Number of unique zip codes:", length(unique(model_data$zipCode)), "\n")

# Split data
set.seed(123)
train_indices <- sample(1:nrow(model_data), 0.8 * nrow(model_data))
train_data <- model_data[train_indices, ]
test_data <- model_data[-train_indices, ]

# Build random forests with different numbers of trees (with corrected categorical zipCode)
rf_1 <- randomForest(SalePrice ~ ., data = train_data, ntree = 1, mtry = 3, seed = 123)
rf_5 <- randomForest(SalePrice ~ ., data = train_data, ntree = 5, mtry = 3, seed = 123)
rf_25 <- randomForest(SalePrice ~ ., data = train_data, ntree = 25, mtry = 3, seed = 123)
rf_100 <- randomForest(SalePrice ~ ., data = train_data, ntree = 100, mtry = 3, seed = 123)
rf_500 <- randomForest(SalePrice ~ ., data = train_data, ntree = 500, mtry = 3, seed = 123)
rf_1000 <- randomForest(SalePrice ~ ., data = train_data, ntree = 1000, mtry = 3, seed = 123)
rf_2000 <- randomForest(SalePrice ~ ., data = train_data, ntree = 2000, mtry = 3, seed = 123)
rf_5000 <- randomForest(SalePrice ~ ., data = train_data, ntree = 5000, mtry = 3, seed = 123)
```




```{r}
#| label: performance-comparison-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Calculate predictions and performance metrics for test data
predictions_1_test <- predict(rf_1, test_data)
predictions_5_test <- predict(rf_5, test_data)
predictions_25_test <- predict(rf_25, test_data)
predictions_100_test <- predict(rf_100, test_data)
predictions_500_test <- predict(rf_500, test_data)
predictions_1000_test <- predict(rf_1000, test_data)
predictions_2000_test <- predict(rf_2000, test_data)
predictions_5000_test <- predict(rf_5000, test_data)

# Calculate predictions for training data
predictions_1_train <- predict(rf_1, train_data)
predictions_5_train <- predict(rf_5, train_data)
predictions_25_train <- predict(rf_25, train_data)
predictions_100_train <- predict(rf_100, train_data)
predictions_500_train <- predict(rf_500, train_data)
predictions_1000_train <- predict(rf_1000, train_data)
predictions_2000_train <- predict(rf_2000, train_data)
predictions_5000_train <- predict(rf_5000, train_data)

# Calculate RMSE for test data
rmse_1_test <- sqrt(mean((test_data$SalePrice - predictions_1_test)^2))
rmse_5_test <- sqrt(mean((test_data$SalePrice - predictions_5_test)^2))
rmse_25_test <- sqrt(mean((test_data$SalePrice - predictions_25_test)^2))
rmse_100_test <- sqrt(mean((test_data$SalePrice - predictions_100_test)^2))
rmse_500_test <- sqrt(mean((test_data$SalePrice - predictions_500_test)^2))
rmse_1000_test <- sqrt(mean((test_data$SalePrice - predictions_1000_test)^2))
rmse_2000_test <- sqrt(mean((test_data$SalePrice - predictions_2000_test)^2))
rmse_5000_test <- sqrt(mean((test_data$SalePrice - predictions_5000_test)^2))

# Calculate RMSE for training data
rmse_1_train <- sqrt(mean((train_data$SalePrice - predictions_1_train)^2))
rmse_5_train <- sqrt(mean((train_data$SalePrice - predictions_5_train)^2))
rmse_25_train <- sqrt(mean((train_data$SalePrice - predictions_25_train)^2))
rmse_100_train <- sqrt(mean((train_data$SalePrice - predictions_100_train)^2))
rmse_500_train <- sqrt(mean((train_data$SalePrice - predictions_500_train)^2))
rmse_1000_train <- sqrt(mean((train_data$SalePrice - predictions_1000_train)^2))
rmse_2000_train <- sqrt(mean((train_data$SalePrice - predictions_2000_train)^2))
rmse_5000_train <- sqrt(mean((train_data$SalePrice - predictions_5000_train)^2))

# Calculate R-squared
r2_1 <- 1 - sum((test_data$SalePrice - predictions_1_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_5 <- 1 - sum((test_data$SalePrice - predictions_5_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_25 <- 1 - sum((test_data$SalePrice - predictions_25_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_100 <- 1 - sum((test_data$SalePrice - predictions_100_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_500 <- 1 - sum((test_data$SalePrice - predictions_500_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_1000 <- 1 - sum((test_data$SalePrice - predictions_1000_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_2000 <- 1 - sum((test_data$SalePrice - predictions_2000_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)
r2_5000 <- 1 - sum((test_data$SalePrice - predictions_5000_test)^2) / sum((test_data$SalePrice - mean(test_data$SalePrice))^2)

# Create performance comparison
performance_df <- data.frame(
  Trees = c(1, 5, 25, 100, 500, 1000, 2000, 5000),
  RMSE_Test = c(rmse_1_test, rmse_5_test, rmse_25_test, rmse_100_test, rmse_500_test, rmse_1000_test, rmse_2000_test, rmse_5000_test),
  RMSE_Train = c(rmse_1_train, rmse_5_train, rmse_25_train, rmse_100_train, rmse_500_train, rmse_1000_train, rmse_2000_train, rmse_5000_train),
  R_squared = c(r2_1, r2_5, r2_25, r2_100, r2_500, r2_1000, r2_2000, r2_5000)
)

print(performance_df)
```



### 1. The Power of More Trees Visualization

- RMSE vs Number of Trees 

The plot shows that the greatest improvement in performance happens upfront, especially as we go from 1 to around 25 trees. Within that range, RMSE falls off sharply, which implies the model is learning meaningful patterns very fast. Beyond this, the curve begins to flatten out, and while adding more trees does continue to lower RMSE, the improvements become much smaller. Once we are into a few hundred trees, we have very minimal gains, and indeed can see clear diminishing returns. In essence, the first few trees give most of the accuracy, while the later trees only finesse the model rather than really improving it.

#### RMSE Plot
```{r}
#| label: rmse-plot
#| echo: false
#| fig-width: 10
#| fig-height: 6

library(ggplot2)
library(tidyr)

# Reshape data to long format for plotting both training and test RMSE
rmse_plot_data <- performance_df %>%
  select(Trees, RMSE_Train, RMSE_Test) %>%
  pivot_longer(cols = c(RMSE_Train, RMSE_Test), 
               names_to = "Dataset", 
               values_to = "RMSE") %>%
  mutate(Dataset = case_when(
    Dataset == "RMSE_Train" ~ "Training",
    Dataset == "RMSE_Test" ~ "Test"
  ))

# Create RMSE plot with log scale on x-axis
ggplot(rmse_plot_data, aes(x = Trees, y = RMSE, color = Dataset)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5) +
  scale_x_log10(breaks = c(1, 5, 25, 100, 500, 1000, 2000, 5000),
                labels = c("1", "5", "25", "100", "500", "1000", "2000", "5000")) +
  labs(
    title = "RMSE vs Number of Trees",
    subtitle = "How RMSE decreases as more trees are added to the random forest",
    x = "Number of Trees (log scale)",
    y = "RMSE",
    color = "Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray50"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

#### R-squared Plot

```{r}
#| label: r-squared-plot
#| echo: false
#| fig-width: 10
#| fig-height: 6

library(ggplot2)

# Create R-squared plot with log scale on x-axis
ggplot(performance_df, aes(x = Trees, y = R_squared)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 2.5, color = "steelblue") +
  scale_x_log10(breaks = c(1, 5, 25, 100, 500, 1000, 2000, 5000),
                labels = c("1", "5", "25", "100", "500", "1000", "2000", "5000")) +
  labs(
    title = "R-squared vs Number of Trees",
    subtitle = "How R-squared increases as more trees are added to the random forest",
    x = "Number of Trees (log scale)",
    y = "R-squared"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray50")
  )
```



### 2. Overfitting Visualization and Analysis



**Your analysis should explain:**


Individual decision trees overfit, where deeper and more complex trees lead to memorization of the training data rather than general patterns. Overfitting is tied to reduced training error but no change-or worsening-test performance. On the other hand, by composing a large number of diverse trees, random forests avoid this problem, which creates a more stable and balanced model. They avoid overfitting through three main mechanisms: bootstrap sampling means each tree sees a different subset of the data; random feature selection at each split allows different trees to consider different predictors; averaging in the final prediction cancels out the noise. These steps ensure the model captures real signal without getting lost in the details of the training data.



Create a side-by-side comparison showing:
1. **Decision Trees:** Training vs Test RMSE as max depth increases (showing overfitting)
2. **Random Forests:** Training vs Test RMSE as number of trees increases (no overfitting)



```{r}
#| label: overfitting-comparison
#| echo: false
#| fig-width: 14
#| fig-height: 6

library(ggplot2)
library(tidyr)
library(rpart)
library(gridExtra)

# Build decision trees with different max depths
max_depths <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 12, 15, 20, 25, 30)
dt_train_rmse <- numeric(length(max_depths))
dt_test_rmse <- numeric(length(max_depths))

for (i in seq_along(max_depths)) {
  dt_model <- rpart(SalePrice ~ ., data = train_data, 
                    control = rpart.control(maxdepth = max_depths[i]))
  
  dt_pred_train <- predict(dt_model, train_data)
  dt_pred_test <- predict(dt_model, test_data)
  
  dt_train_rmse[i] <- sqrt(mean((train_data$SalePrice - dt_pred_train)^2))
  dt_test_rmse[i] <- sqrt(mean((test_data$SalePrice - dt_pred_test)^2))
}

# Create decision tree data frame
dt_performance <- data.frame(
  MaxDepth = max_depths,
  RMSE_Train = dt_train_rmse,
  RMSE_Test = dt_test_rmse
)

# Reshape decision tree data for plotting
dt_plot_data <- dt_performance %>%
  pivot_longer(cols = c(RMSE_Train, RMSE_Test),
               names_to = "Dataset",
               values_to = "RMSE") %>%
  mutate(Dataset = case_when(
    Dataset == "RMSE_Train" ~ "Training",
    Dataset == "RMSE_Test" ~ "Test"
  ))

# Reshape random forest data for plotting (using existing performance_df)
rf_plot_data <- performance_df %>%
  select(Trees, RMSE_Train, RMSE_Test) %>%
  pivot_longer(cols = c(RMSE_Train, RMSE_Test),
               names_to = "Dataset",
               values_to = "RMSE") %>%
  mutate(Dataset = case_when(
    Dataset == "RMSE_Train" ~ "Training",
    Dataset == "RMSE_Test" ~ "Test"
  ))

# Calculate common y-axis limits for both plots
y_min <- min(c(dt_plot_data$RMSE, rf_plot_data$RMSE))
y_max <- max(c(dt_plot_data$RMSE, rf_plot_data$RMSE))
y_range <- c(y_min, y_max)

# Create decision tree plot
p1 <- ggplot(dt_plot_data, aes(x = MaxDepth, y = RMSE, color = Dataset)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(limits = y_range) +
  labs(
    title = "Decision Trees: Overfitting with Complexity",
    subtitle = "Training RMSE decreases while test RMSE increases (overfitting)",
    x = "Max Depth (Tree Complexity)",
    y = "RMSE",
    color = "Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )

# Create random forest plot
p2 <- ggplot(rf_plot_data, aes(x = Trees, y = RMSE, color = Dataset)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_x_log10(breaks = c(1, 5, 25, 100, 500, 1000, 2000, 5000),
                labels = c("1", "5", "25", "100", "500", "1000", "2000", "5000")) +
  scale_y_continuous(limits = y_range) +
  labs(
    title = "Random Forests: No Overfitting",
    subtitle = "Both training and test RMSE decrease together",
    x = "Number of Trees (log scale)",
    y = "RMSE",
    color = "Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )

# Display plots side by side
grid.arrange(p1, p2, ncol = 2)
```

### 3. Linear Regression vs Random Forest Comparison

```{r}
#| label: linear-regression-comparison
#| echo: false

# Build linear regression model
lm_model <- lm(SalePrice ~ ., data = train_data)

# Calculate predictions and RMSE for linear regression
lm_pred_test <- predict(lm_model, test_data)
lm_rmse_test <- sqrt(mean((test_data$SalePrice - lm_pred_test)^2))

# Get RMSE values for random forests (already calculated)
rf_1_rmse <- rmse_1_test
rf_100_rmse <- rmse_100_test
rf_1000_rmse <- rmse_1000_test

# Calculate percentage improvements over linear regression
# Percentage improvement = ((baseline - new) / baseline) * 100
pct_improvement_1 <- ((lm_rmse_test - rf_1_rmse) / lm_rmse_test) * 100
pct_improvement_100 <- ((lm_rmse_test - rf_100_rmse) / lm_rmse_test) * 100
pct_improvement_1000 <- ((lm_rmse_test - rf_1000_rmse) / lm_rmse_test) * 100

# Create comparison table
comparison_table <- data.frame(
  Model = c("Linear Regression", 
            "Random Forest (1 tree)", 
            "Random Forest (100 trees)", 
            "Random Forest (1000 trees)"),
  RMSE_Test = c(lm_rmse_test, rf_1_rmse, rf_100_rmse, rf_1000_rmse),
  Improvement = c("Baseline", 
                  paste0(round(pct_improvement_1, 2), "%"),
                  paste0(round(pct_improvement_100, 2), "%"),
                  paste0(round(pct_improvement_1000, 2), "%"))
)

# Format and display table
knitr::kable(comparison_table,
             caption = "Model Comparison: Linear Regression vs Random Forests",
             col.names = c("Model", "Test RMSE", "Improvement over Linear Regression"),
             digits = 2,
             align = c("l", "r", "r"))
```

**Analysis of Results:**

The comparison table reveals several important insights about model performance:

**Improvement from 1 tree to 100 trees:** The transition from a single decision tree (Random Forest with 1 tree) to 100 trees is a significant improvement in RMSE. This is a great illustration of the "wisdom of the crowd," a phenomenon that occurs when multiple weak models (single trees, in this case) are used to produce a far stronger model (the Random Forest). The reason for this improvement is that there is a convergence effect when a forecast from a single tree is combined with a forecast from a second, different tree, resulting in a more accurate prediction.

**Linear Regression vs 100-tree Random Forest:** Moreover, the gain from switching from linear regression to a random forest of 100 trees is large and often comparable or larger than that from 1 to 100 trees. From this, it seems reasonable to conclude that random forests capture nonlinear relationships and interactions among the variables that liner regression does not catch, which again is highly useful when working with complex datasets containing nonlinear patterns.

**When Random Forests are Worth the Complexity:** Random forests are worth the added complexity when:

1. The relationship between predictors and the target variable is non-linear  
2. There are important interactions between variables that linear models miss
3. The dataset is large enough to support the computational cost
4. Prediction accuracy is more important than model interpretability
5. The performance gain (as shown in the RMSE reduction) justifies the increased computational time and reduced interpretability
 

**Trade-offs between Interpretability and Performance:** Linear regression provides a high level of interpretability because it is easy to interpret how every variable, including the dependent variable, contributes to a particular result because of the use of coefficients. Random forests, which are more accurate, are considered "black box" models because it is hard to interpret how a particular prediction result is obtained. It is, therefore, easy to see that the trade-off here is that accuracy is gained at the cost of interpretability. In most cases, especially when interpretability is paramount, such that models are used to address legal obligations, scientific presentations, and so on, linear regression models might still be used, even though they are less accurate.